*Symbiodinium* communities in reef corals of St. John, U.S.V.I. (2012)

### 2016-08-27 analysis/STJ2012_analysis.Rmd
begin pairwiseAlignment analysis

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 3, digits = 2)
```

@@

# Data pre-processing
### Import dataset
The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object (class 'phyloseq') to facilitate downstream analyses.
```{r data_import, cache=T}
# Import OTU tables
OTU97 <- otu_table(read.table("data/97_otus.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
OTU97bs <- otu_table(read.table("data/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
OTU97bsp <- otu_table(read.table("data/97_otus_byspecies.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
OTU100 <- otu_table(read.table("data/100_otus.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy
TAX97 <- tax_table(as.matrix(read.table("data/97_otus_tax_assignments.txt", sep="\t", row.names=2,
                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
TAX97bs <- tax_table(as.matrix(read.table("data/97_otus_bysample_tax_assignments.txt", sep="\t", row.names=2,
                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
TAX97bsp <- tax_table(as.matrix(read.table("data/97_otus_byspecies_tax_assignments.txt", sep="\t", row.names=2,
                                        col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
TAX100 <- tax_table(as.matrix(read.table("data/100_otus_tax_assignments.txt", sep="\t", row.names=2,
                                         col.names=c("Clade", "OTU.ID", "Taxonomy", "Eval", "Subtype"))))
# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
# Build phyloseq objects
phy97 <- phyloseq(OTU97, TAX97, SAM)
phy97bs <- phyloseq(OTU97bs, TAX97bs, SAM)
phy97bsp <- phyloseq(OTU97bsp, TAX97bsp, SAM)
phy100 <- phyloseq(OTU100, TAX100, SAM)
```

### Filter dataset
```{r filter_dataset, cache=T}
# Filter OTUs by minimum count
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97)[which(taxa_sums(phy97) >= n)]
taxa97bs <- taxa_sums(phy97bs)[which(taxa_sums(phy97bs) >= n)]
taxa97bsp <- taxa_sums(phy97bsp)[which(taxa_sums(phy97bsp) >= n)] 
taxa100 <- taxa_sums(phy100)[which(taxa_sums(phy100) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs)
phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp)
phy100.f <- prune_taxa(names(taxa100), phy100)

@@

## Background symbionts
#### (<1% relative abundance)
```{r bk.net, fig.width=5, fig.height=5, fig.align="center"}
# filter out OTUs that only occurred in one sample
phy97bs.f.p2 <- filter_taxa(phy97bs.f.p, function(x) sum(x>0) > 1, prune=TRUE)
# Get network with one node for each species.
otudf <- data.frame(t(otu_table(phy97bs.f.p2)), check.names=F)
# Percent of samples of species that contain OTU at less than 1% relative abundance (but more than 0%)

sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p2))$Species), 
                   FUN=function(x) any(x>0 & x<0.01))
rownames(sp.ag) <- sp.ag$Species
sp.ag <- sp.ag[, -1]
edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
edges <- droplevels(edges[edges$weight>0, ])
# Create "nodes" table from sample and tax data
sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p2))$Species))
tdf <- data.frame(tax_table(phy97bs.f.p2))[rownames(data.frame(tax_table(phy97bs.f.p2))) %in% edges$target, ]
tdf$id <- rownames(tdf)
nodes <- merge(sdf, tdf, all=T)
nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
# Create network
net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 3, digits = 9)
```

@@

# Data pre-processing
### Import dataset
The [phyloseq](https://joey711.github.io/phyloseq/) package is used here to combine the OTU table, taxonomic assignments, and sample metadata into a single R data object (class 'phyloseq') to facilitate downstream analyses.
```{r data_import, cache=T}

```{r filter_taxa}
# Import taxonomic assignment data from local and global alignments
taxcolnames <- c("OTU", "BLAST.hit", "BLAST.pct.id", "BLAST.align.length", "BLAST.mismatches", "BLAST.gapopen", 
                 "BLAST.q.start", "BLAST.q.end", "BLAST.s.start", "BLAST.s.end", "BLAST.e.val", "BLAST.bit.score", 
                 "needle.hit", "needle.length", "needle.score")

TAX97 <- read.table("data/otus_97/tax_results.txt", sep=" ", col.names=taxcolnames)
TAX97$needle.score <- as.numeric(gsub("[^0-9.]", "", TAX97$needle.score))

TAX97bs <- read.table("data/otus_97_bysample/tax_results.txt", sep=" ", col.names=taxcolnames)
TAX97bs$needle.score <- as.numeric(gsub("[^0-9.]", "", TAX97bs$needle.score))

TAX97bsp <- read.table("data/otus_97_byspecies/tax_results.txt", sep=" ", col.names=taxcolnames)
TAX97bsp$needle.score <- as.numeric(gsub("[^0-9.]", "", TAX97bsp$needle.score))

TAX100 <- read.table("data/otus_100/tax_results.txt", sep=" ", col.names=taxcolnames)
TAX100$needle.score <- as.numeric(gsub("[^0-9.]", "", TAX100$needle.score))

# Merge identical taxonomic assignments (because some reference sequences are not unique...)
ident <- readLines("data/ITS2db_trimmed_notuniques_otus.txt")
ident <- strsplit(ident, split="\t")
names(ident) <- c("C3_multiple", "B19_multiple", "H1_multiple", "C33_multiple", "H5_multiple", "A2_multiple",
                  "B21/B38_multiple", "C35/C26_multiple", "A1/A13_multiple", "F5.2_multiple", "D1_multiple", 
                  "A10_multiple", "B2_multiple", "A1.2_multiple", "C1c/C45_multiple", "B1_multiple",
                  "C91_multiple", "C82_multiple", "D1a_multiple", "A4a_multiple", "C1_multiple",
                  "F5.2d_multiple", "C21/C32_multiple", "A3/A6_multiple", "C26/C35a_multiple")

merge.identical.hits <- function(x) {
  sapply(as.character(x), FUN=function(y) {
    ifelse(y %in% unlist(ident), 
           gsub("[0-9]*$", "", names(unlist(ident))[match(y, unlist(ident))]), 
           y)
    })
}

TAX97$BLAST.matched <- merge.identical.hits(TAX97$BLAST.hit)
TAX97$needle.matched <- merge.identical.hits(TAX97$needle.hit)

# OTUs for which BLAST and needle gave different assignments
nomatch <- subset(TAX97, BLAST.matched!=needle.matched)
nomatch

# Get percent identity of global alignment to needle hits
# import reference sequences
ref <- readDNAStringSet("data/ITS2db_trimmed.fasta")
# import otu sequences
otus97 <- readDNAStringSet("data/otus_97/97_otus_rep_set.fasta")
names(otus97) <- gsub(" .*$", "", names(otus97))
# align each otu sequence to its best needle hit, using same parameters as command line needle
# "global-local" forces whole reference/pattern sequence to be used, but subject/query can be a concescutive subsequence -- this is because queries may include slightly longer region of gene than references because references were trimmed by o-smart-trim. this means that query sequence can have more bases than the reference at beginning or end of the sequence (=end gaps in alignment) without penalties.
EDNAFULL <- as.matrix(read.table("data/EDNAFULL", skip=8, header=T, row.names=1))
n.psa <- pairwiseAlignment(subject=otus97[TAX97$OTU], pattern=ref[TAX97$needle.hit], type="global-local",
                           substitutionMatrix=EDNAFULL,
                           gapOpening=10.0, gapExtension=0.5)

# calculate percent similarity to reference with indels counting as single difference regardless of length
TAX97 <- within(TAX97, {
  needle.hit.length <- width(pattern(n.psa))
  needle.mismatches <- nmismatch(n.psa)
  needle.insertions <- unlist(lapply(as.list(insertion(n.psa)), length))
  needle.deletions <- unlist(lapply(as.list(deletion(n.psa)), length))
  needle.similarity <- (needle.hit.length-needle.mismatches-needle.insertions-needle.deletions)/needle.hit.length
})

g.psa <- pairwiseAlignment(subject=otus97[TAX97$OTU], pattern=ref[TAX97$needle.hit], type="global",
                           substitutionMatrix=EDNAFULL,
                           gapOpening=10.0, gapExtension=0.5)

# calculate percent similarity to reference with indels counting as single difference regardless of length
TAX97 <- within(TAX97, {
  gneedle.hit.length <- width(pattern(g.psa))
  gneedle.mismatches <- nmismatch(g.psa)
  gneedle.insertions <- unlist(lapply(as.list(insertion(g.psa)), length))
  gneedle.deletions <- unlist(lapply(as.list(deletion(g.psa)), length))
  gneedle.similarity <- (gneedle.hit.length-gneedle.mismatches-gneedle.insertions-gneedle.deletions)/gneedle.hit.length
})

aligned(n.psa[])
n.psa
aligned(n.psa[1])
TAX97$needle.hit.length <- width(pattern(n.psa))
TAX97$needle.mismatches <- nmismatch(n.psa)
TAX97$needle.insertions <- unlist(lapply(as.list(insertion(n.psa)), length))
TAX97$needle.deletions <- unlist(lapply(as.list(deletion(n.psa)), length))
TAX97$needle.similarity <- with(TAX97, (needle.hit.length - needle.mismatches - needle.insertions - needle.deletions) / needle.hit.length)

with(TAX97[115, ], {
  aln <<- pairwiseAlignment(subject=otus97[OTU], pattern=ref[needle.hit], type="global-local",
                            substitutionMatrix=EDNAFULL,
                            gapOpening=10.0, gapExtension=0.5)
})
aln

aln <- pairwiseAlignment(subject="ACTTCTCTATCTATCTTGGTACG", pattern="AATTCTCGGTACG", type="global")
aln2 <- pairwiseAlignment(subject="ACTTCTACG", pattern="AATTCTCGGTACG", type="global")
aln
aln2
width(aligned(aln))
nmatch(aln)/width(aligned(aln))

# calculate pct id of all positions relative to reference/(pattern)
nmatch(aln)/width(pattern(aln))
# calculate pct id counting indels as single difference regardless of length = GAST definition
(width(pattern(aln)) - nmismatch(aln) - length(unlist(insertion(aln))) - length(unlist(deletion(aln))))/width(pattern(aln))

stringDist(aln)

nmatch(aln)
nmismatch(aln)
width(nindel(aln))
length(unlist(insertion(aln))); length(unlist(deletion(aln)))
length(unlist(insertion(aln)))
length(unlist(deletion(aln)))
insertion(aln)
deletion(aln)
unlist(nindel(aln))
nindel(aln)$insertion
str(nindel(aln))
b <- n.psa[11]
as.list(indel(subject(b)))
as.list(indel(subject(aln)))
length(as.data.frame(deletion(b))[[1]])
indel(b)
length(unlist(insertion(b)))

class(EDNAFULL)
n.psa[47]
summary(n.psa[47])
nmatch(n.psa[47])
nindel(n.psa[47])
nmismatch(n.psa[47])
nchar(pattern(n.psa[2]))
246/255
nmismatch(n.psa[2])

```{r build_phyloseq}
# Import OTU tables
OTU97 <- otu_table(read.table("data/otus_97/97_otus.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
OTU97bs <- otu_table(read.table("data/otus_97_bysample/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
#OTU97bsp <- otu_table(read.table("data/97_otus_byspecies.tsv", header=T, check.names=F, row.names=1,
#                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
OTU100 <- otu_table(read.table("data/otus_100/100_otus.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
# Import taxonomy

# Import sample data
SAM <- sample_data(read.delim("data/mapping_file.txt", sep="\t", header=T, row.names=1))
# Build phyloseq objects
phy97 <- phyloseq(OTU97, TAX97, SAM)
phy97bs <- phyloseq(OTU97bs, TAX97bs, SAM)
phy97bsp <- phyloseq(OTU97bsp, TAX97bsp, SAM)
phy100 <- phyloseq(OTU100, TAX100, SAM)
```

### Filter dataset
```{r filter_dataset, cache=T}

# Filter OTUs by minimum count
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97)[which(taxa_sums(phy97) >= n)]
taxa97bs <- taxa_sums(phy97bs)[which(taxa_sums(phy97bs) >= n)]
taxa97bsp <- taxa_sums(phy97bsp)[which(taxa_sums(phy97bsp) >= n)] 
taxa100 <- taxa_sums(phy100)[which(taxa_sums(phy100) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs)
phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp)
phy100.f <- prune_taxa(names(taxa100), phy100)

@@

## Background symbionts
#### (<1% relative abundance)
```{r bk.net, fig.width=5, fig.height=5, fig.align="center"}
# filter out OTUs that only occurred in one sample
phy97bs.f.p2 <- filter_taxa(phy97bs.f.p, function(x) sum(x>0) > 1, prune=TRUE)
# Get network with one node for each species.
otudf <- data.frame(t(otu_table(phy97bs.f.p2)), check.names=F)
# Whether any samples of species contain OTU at less than 1% relative abundance (but more than 0%)
sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p2))$Species), 
                   FUN=function(x) any(x>0 & x<0.01))
rownames(sp.ag) <- sp.ag$Species
sp.ag <- sp.ag[, -1]
edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
edges <- droplevels(edges[edges$weight>0, ])  # only keep if was observed in more than N colony of a spp
# Create "nodes" table from sample and tax data
sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p2))$Species))
tdf <- data.frame(tax_table(phy97bs.f.p2))[rownames(data.frame(tax_table(phy97bs.f.p2))) %in% edges$target, ]
tdf$id <- rownames(tdf)
nodes <- merge(sdf, tdf, all=T)
nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
# Create network
net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

@@

set.seed(42399)
l3 <- layout_nicely(net3)
l3 <- norm_coords(l3, ymin=-1, ymax=1, xmin=-1, xmax=1)
par(mar=c(0,0,0,0))
V(net3)$size <- ifelse(is.na(V(net3)$Clade), 15, 10*degree(net3)^0.2)
plot(net3, rescale=F, layout=l3*1, edge.curved=0.1, vertex.label.cex=V(net3)$size/20,
     vertex.label.color="black")
